AFQ-Insight

### 2019-12-16 reports/manuscript/sections/introduction.tex
Use tract and bundle consistently. Add sentence to introduction arguing that bundles, not streamlines, are the appropriate degrees of freedom for studying dMRI

To relate the diffusion in each voxel to the macro-structure of long-range
connections between different brain regions, methods for computational
tract-tracing from diffusion MRI, or tractography, combine the estimates of
fascicle orientations in each voxel to form streamlines that traverse the volume
of the white matter \cite{Conturo1999-je, Mori2002-qi}. These methods have been
under increased scrutiny and several lines of investigation have raised
critiques of their validity \cite{Maier-Hein2017-vb, Thomas2014-ki}. On the
other hand, there have been efforts to shore up the inferences made with these
methods \cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel nerve fiber
bundles requires extraordinary evidence, and delineating the exact cortical
termination of the streamlines in the gray matter is still prone to error, there
is little dispute that tractography can accurately define the location of
several major white matter bundles that are known to exist within the core of
the white matter \cite{Maier-Hein2017-vb}.

Leveraging this fact, one of the most powerful methods currently
available to put macro- and micro-structure together is
\emph{tractometry}: assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are
several different available implementations of this overall idea,
the principles are similar \cite{yeatman2012tract, Yendiki2011-ay,
Wassermann2016-iv, ODonnell2009-uu}: tractometry begins by delineating
the parts of the white matter that belong to different major tracts
(i.e. bundles of streamlines), such as the corticospinal tract or arcuate
fasciculus, and samples the biophysical properties (such as fractional
anisotropy or mean diffusivity) along the length of these bundles.
In some cases, tissue properties along the length of each bundle are
summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability along
the trajectory of each bundle. This justifies retaining the individual
samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu}. However, while this retains important
information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals
to each other. This is usually done following one of the following
approaches:

>>>>

To relate the diffusion in each voxel to the macro-structure of
long-range connections between different brain regions, methods for
computational tract-tracing from diffusion MRI, or tractography, combine
the estimates of fascicle orientations in each voxel to form streamlines
that traverse the volume of the white matter \cite{Conturo1999-je,
Mori2002-qi}. These methods have been under increased scrutiny and
several lines of investigation have raised critiques of their validity
\cite{Maier-Hein2017-vb, Thomas2014-ki}. On the other hand, there
have been efforts to shore up the inferences made with these methods
\cite{Pestilli2014NatMeth, Takemura2016-sh, Smith2013-nc, Smith2015-cx,
Smith2015-zt, Rheault2018-wk}. Importantly, though discovering novel
nerve fiber tracts requires extraordinary evidence, and delineating
the exact cortical termination of the streamlines in the gray matter
is still prone to error, there is little dispute that tractography
can accurately define the location of several major white matter
tracts that are known to exist within the core of the white matter
\cite{Maier-Hein2017-vb}.

Leveraging this fact, one of the most powerful methods currently
available to put macro- and micro-structure together is
\emph{tractometry}: assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are
several different available implementations of this overall idea,
the principles are similar \cite{yeatman2012tract, Yendiki2011-ay,
Wassermann2016-iv, ODonnell2009-uu}: tractometry begins by delineating
the parts of the white matter that belong to different major ``tracts''
(i.e. anatomical or functional groups of white matter fibers),
such as the corticospinal tract or arcuate fasciculus, assigning
tractography generated streamlines to ``bundles,'' which approximate
the anatomical tracts, and sampling biophysical properties (such
as fractional anisotropy or mean diffusivity) along the length
of these bundles. Whereas the identification of the anatomical
correlates of tractography-generated streamlines is an open problem,
the correspondence between tractometry's bundles and anatomical tracts
is well supported. Tract-based diffusion metrics are therefore
appropriate degrees of freedom with which to understand the
white-matter correlates of cognitive function or disease.

In some previous tractometry-based studies, tissue properties along
the length of each tract were summarized by taking the mean along
each bundle, but there is a large body of evidence showing that there
is systematic variability along the trajectory of each bundle. This
justifies retaining the individual samples along the length of each
bundle \cite{yeatman2012tract, colby2012, ODonnell2009-uu}. While this
retains important information about each individual's white matter, it
also presents statistical challenges due to the dimensionality of the
data. Based on tractometry, researchers may choose to compare different
individuals to each other. This is usually done according to one of the
following approaches:

### 2019-12-16 reports/manuscript/sections/methods.tex
Use tract and bundle consistently. Add sentence to introduction arguing that bundles, not streamlines, are the appropriate degrees of freedom for studying dMRI

Data from both of these studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox (AFQ)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Major tracts were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a tract profile of tissue properties in each tract was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
tract. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalnobis distance of the node from the average of that node across
streamlines. This means that streamlines that are more representative of
the tract contribute more to the tract profile, relative to streamlines
that are on the edge of the tract.

>>>>

Data from both of these studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox (AFQ)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Major tracts were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a tract profile of tissue properties in each bundle was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
tract. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalnobis distance of the node from the average of that node across
streamlines. This means that streamlines that are more representative of
the tract contribute more to the tract profile, relative to streamlines
that are on the edge of the tract.
